### 2014-08-14 doc/thesis/typedlua.tex
Adding text about the refinement of table types

The declaration of the interface \texttt{Person} is an alias to the table type
$\{``firstname":\String, ``lastname":\String\}$ in the remainder
of the current scope.
Now we can use \texttt{Person} in type declarations:
\begin{verbatim}
    local function greeter (person:Person):string
      return "Hello " .. person.firstname ..
             " " .. person.lastname
    end

@@

    print(greeter(user1)) -- Hello, Lewis Allan
    print(greeter(user2)) -- compile-time error
    print(greeter(user3)) -- Hello, Lou Reed
    print(greeter(user4)) -- compile-time error
\end{verbatim}

@@

>>>>

The declaration of the interface \texttt{Person} is an alias to the table type
$\{``firstname":\String, ``lastname":\String\}$ in the remainder
of the current scope.
Now we can use \texttt{Person} in type declarations:
\begin{verbatim}
    local function byebye (person:Person):string
      return "Goodbye " .. person.firstname ..
             " " .. person.lastname
    end

@@

    print(greeter(user1)) -- Goodbye Lewis Allan
    print(greeter(user2)) -- compile-time error
    print(greeter(user3)) -- Goodbye Lou Reed
    print(greeter(user4)) -- compile-time error
\end{verbatim}

@@

Lua programmers often build records incrementally, that is,
they usually declares a local variable with an empty table,
and then use assignment to add fields to this table:
\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
    print("bye bye " .. person.firstname)  -- bye bye Lou
\end{verbatim}

In this example, we want to refine the type of \texttt{person} as
we build the table:
starting with $\{\}$, and then refining to $\{``firstname":\String\}$,
and finally reaching $\{``firstname":\String, ``lastname":\String\}$.
This type change is trickier than the one that we introduced for narrowing
union types, as we are not just allowing the programmer to change
the type of the variable \texttt{person}, but we are actually allowing
the programmer to change the type of the value that \texttt{person} points to.
This type change is safe in this example, but it is not in the following one:
\begin{verbatim}
    local person = {}
    local bogus = person
    person.firstname = "Lou"
    person.lastname = "Reed"
    bogus.firstname = true               -- compile-time error
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, the fifth line is unsound because it allows the
programmer to change the type of the value that is stored in
\texttt{person.firstname}, and makes the last line crash during
run-time.
On the first line, we assign an empty table to \texttt{person}.
On the second line, we assign the type of \texttt{person} to
\texttt{bogus}.
On the third and fourth lines we change the type of \texttt{person}
as we did in the previous example.
On the fifth line we try to change the type of \texttt{bogus}
from $\{\}$ to $\{``firstname":\Boolean\}$, but if we allow this
type change we also allow changing the value that is stored in
\texttt{person.firstname}, regardless of its type.
Taking the changes individually looks fine, but the truth is that
aliasing makes one of them unsound.

The location of the change also matters, as the next example shows:
\begin{verbatim}
    local person = { lastname = "Reed" }
    local function spoil ()
      person.firstname = nil             -- compile-time error
    end
    person.firstname = "Lou"
    spoil()
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, the third line is unsound because it allows the
programmer to change the type of \texttt{person} outside of the scope
that it was declared.
The call to \texttt{spoil} erases the field \texttt{firstname} from
\texttt{person}, and makes the last line crash during run-time.

It is also not sound to allow the type of the fields to change,
as the next example shows:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1964              -- compile-time error
    person.lastname = 2013               -- compile-time error
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, both third and fourth line are unsound because
they are trying to change the type of fields that already exist.

Typed Lua tags a variable that holds a table type as either
\emph{open} or \emph{closed}.
If a variable gets its type from a table constructor then it
is open, otherwise it is always closed.

Typed Lua uses three restrictions to decide whether it is
sound to allow field assignment to change the type of an
open variable or not:
the variable must be local to the current block,
the new type must be a subtype of the old one, and
the variable cannot have been assigned in another function.

We can create aliases to open variables, but they are always closed.
Any mutation on these aliases is not a problem, as the type of
the original reference can only change towards a subtype.
For mutable fields this means that the type of the field cannot
change after it is added to the type of the table.

Table types along with their refinement through field assignment
are the foundation for defining modules and objects in Typed Lua.
In the next section, we show that the refinement of table types
through field assignment is crucial for making Typed Lua
support idiomatic definition of Lua modules.
